#VIF Analysis 

#Set WD and install packages
getwd()
install.packages("usdm")
library(usdm)
install.packages("phylolm")
library(phylolm)
Ungulate <- read.csv("UngulateTraits.csv", delim = 'x', header = TRUE)

#VIF analysis

title: "VIF_analysis_Ungulate_Traits"
output: html_document
---
  
  ``{r setup include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
{r}``
#load required package
library(usdm)
#general code for VIF analysis
vif(Ungulate[,c(8:59)])
``

``{r}
#load variable overview
project_root <- 'C:/Users/Kes Lippert/Desktop/Domesticability R project/Domesticability-Analysis-2/'
variable_file <- paste(c(project_root, 'GIS/Variables_overview.md'), collapse = '')
variable_overview <- read.delim(variable_file)
variable_overview <- variable_overview[1:52,1:2]

## Hypothesis

The following variables are likely to be critical for the analysis (and thus best be not deleted): 
1. 
2. 

## Consecutive rounds of the VIF analysis 

Do consecutive rounds of the VIF analysis, while leaving out one variable each round. Guidelines:
1. In the end, the VIF value has to be lower than 10 for each variable, so variables have to be deleted until this is the case. 
2. Preferably delete the one with the highest VIF value each round, this will speed up the decrease of all the VIF values, but:  
3. Make sure that it makes sense biologically and makes sense in the context of the specific trait. Especially tread lightly around the variables we selected previously. 

### VIF-analysis purely based on highest VIF value
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), 
#BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), 
#PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), 
vif(GIS_woody[,c(3:5, 14, 15, 17, 18, 20, 22:24, 38, 39, 42, 43)])
#result is 3 left of the variables we wanted. 
```

### Round one
```{r}
#all variables
vif(GIS_woody[,c(3:43)])
```
BIO5, BIO6 and BIO7 all are infinite. 
  BIO5 = Max Temperature of Warmest Month
  BIO6 = Min Temperature of Coldest Month
  BIO7 = Temperature Annual Range (BIO5-BIO6)
BIO7 is a combination of BIO5 and BIO6, which explains the correlated VIF value. Best step would be to exclude BIO7 from the analysis, according to the hypothesis, an annual temperature range would not tell us much. 

### Round two
```{r}
#excluding: BIO7 (column 31)
vif(GIS_woody[,c(3:30, 32:43)])
```
BIO11 has the highest value. 
  BIO11 = Mean Temperature of Coldest Quarter
It is okay to leave out this variable, the lowest temperature of the coldest quarter would be more informative (as frost temperatures might have an influence).

### Round three
```{r}
#excluding: BIO7 (31), BIO11 (35)
vif(GIS_woody[,c(3:30, 32:34, 36:43)]) 
```
Continentality has the highest VIF value.
  continentality = Average temp. of warmest and coldest month
It would be okay to leave out this variable, as the minimal temperature is enough to say something about the presence/absence of frost. And this does not tell us anything about drought directly. 

### Round four
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9)
vif(GIS_woody[,c(3:8, 10:30, 32:34, 36:43)]) 
```
BIO10 has the highest VIF value
  BIO10 = Mean Temperature of Warmest Quarter
It is okay to leave this value out, there are still other variables included that can give information on the temperature.

### Round five
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34)
vif(GIS_woody[,c(3:8, 10:30, 32, 33, 36:43)]) 
```
maxTempColdestMonth and bio4 have the two highest VIF values.
  maxTempColdestMonth = Maximum temp. of the coldest month
  BIO4 = Temperature Seasonality (standard deviation *100)
    Google: The conventional definition of temperature seasonality is the difference between the annual maximum and minimum temperatures
    Bio4: The amount of temperature variation over a given year (or averaged years) based on the standard deviation (variation) of monthly temperature 
    averages
Both are not directly important for this trait. I expect BIO4 to have a high correlation with other variables, so I'll exclude BIO4. 
### Round six
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28)
vif(GIS_woody[,c(3:8, 10:27, 29, 30, 32, 33, 36:43)]) 
```
BIO1 has the highest VIF value.
BIO1 = Annual Mean Temperature
Not directly important according to the hypothesis, so okay to leave out.
### Round seven
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25)
vif(GIS_woody[,c(3:8, 10:24, 26, 27, 29, 30, 32, 33, 36:43)]) 
```
growingDegDays0 has the highest value.
growingDegDays0 = Sum of months with temperatures greater than 0 degrees
This is important for the analysis. If I delete this variable, I have to make sure that BIO6 stays in the analysis, to account for the frost related hypothesis.
### Round eight
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11)
vif(GIS_woody[,c(3:8, 10, 12:24, 26, 27, 29, 30, 32, 33, 36:43)]) 
```
maxTempColdestMonth has the highest VIF value.
maxTempColdestMonth = Maximum temp. of the coldest month
Not directly important for this analysis, so it can be excluded.
### Round nine
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13),
vif(GIS_woody[,c(3:8, 10, 12, 14:24, 26, 27, 29, 30, 32, 33, 36:43)]) 
```
PETColdestQuarter has the highest VIF value, followed by annualPET. 
PETColdestQuarter = Mean monthly PET of coldest quarter
annualPET = Annual potential evapotranspiration
This indicates that we won't be able to keep all the PET variables. Let's try to keep at least one set. 
### Round ten
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16)
vif(GIS_woody[,c(3:8, 10, 12, 14, 15, 17:24, 26, 27, 29, 30, 32, 33, 36:43)]) 
```
BIO16 has the highest VIF value.
BIO16 = Precipitation of Wettest Quarter
This can be excluded, there are still sets.
### Round eleven
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40)
vif(GIS_woody[,c(3:8, 10, 12, 14, 15, 17:24, 26, 27, 29, 30, 32, 33, 36:39, 41:43)]) 
```
annualPET has the highest VIF value.
annualPET = Annual potential evapotranspiration
This can be excluded, there are still sets. 
### Round twelve
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6)
vif(GIS_woody[,c(3:5, 7, 8, 10, 12, 14, 15, 17:24, 26, 27, 29, 30, 32, 33, 36:39, 41:43)]) 
```
BIO12 has the highest vIF value.
BIO12 = Annual Precipitation
This can be excluded (same as annualPET), there are still sets.
### Round thirteen
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36)
vif(GIS_woody[,c(3:5, 7, 8, 10, 12, 14, 15, 17:24, 26, 27, 29, 30, 32, 33, 37:39, 41:43)]) 
```
BIO17 has the highest VIF value.
BIO17 = Precipitation of Driest Quarter
This is the other part of BIO16, so okay to exclude.
### Round fourteen
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41)
vif(GIS_woody[,c(3:5, 7, 8, 10, 12, 14, 15, 17:24, 26, 27, 29, 30, 32, 33, 37:39, 42, 43)]) 
```
thermicityIndex has the highest VIF value.
thermIndex = Compensated thermicity index
Temperature related, not specifically important for this trait. So okay to delete.
### Round fifteen
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21)
vif(GIS_woody[,c(3:5, 7, 8, 10, 12, 14, 15, 17:20, 22:24, 26, 27, 29, 30, 32, 33, 37:39, 42, 43)]) 
```
BIO5 has the highest VIF value.
BIO5 = Max Temperature of Warmest Month
Temperature related, not specifically important for this trait. So okay to delete.
### Round sixteen
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29)
vif(GIS_woody[,c(3:5, 7, 8, 10, 12, 14, 15, 17:20, 22:24, 26, 27, 30, 32, 33, 37:39, 42, 43)]) 
```
BIO6 has the highest VIF value
BIO6 = Min Temperature of Coldest Month
Is okay to exclude for now (related to BIO5). Will try to add it back in the end and see what happens. 
### Round seventeen
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30)
vif(GIS_woody[,c(3:5, 7, 8, 10, 12, 14, 15, 17:20, 22:24, 26, 27, 32, 33, 37:39, 42, 43)]) 
```
growingDegDays5 has the highest VIF value.
growingDegDays5 = Sum of months with temperatures greater than 5 degrees
Temperature related, not specifically important for this trait. So okay to delete.
### Round eightteen
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12)
vif(GIS_woody[,c(3:5, 7, 8, 10, 14, 15, 17:20, 22:24, 26, 27, 32, 33, 37:39, 42, 43)]) 
```
climaticMoistureIndex has the highest VIF value. Followed by PETWarmestQuarter
climaticMoistureIndex = Metric of relative wetness and aridity
PETWarmestQuarter = Mean monthly PET of warmest quarter
ClimaticMoistureIndex has a lot of information, when looking at the hypothesis. So delete PETWarmestQuarter.
### Round nineteen
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), PETWarmestQuarter (19)
vif(GIS_woody[,c(3:5, 7, 8, 10, 14, 15, 17, 18, 20, 22:24, 26, 27, 32, 33, 37:39, 42, 43)]) 
```
climaticMoistureIndex has the highest VIF value. Followed by BIO8. 
climaticMoistureIndex = Metric of relative wetness and aridity
BIO8 = Mean Temperature of Wettest Quarter
ClimaticMoistureIndex has a lot of information, when looking at the hypothesis. And BIO8 is not important, so delete BIO8.
### Round twenty
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32)
vif(GIS_woody[,c(3:5, 7, 8, 10, 14, 15, 17, 18, 20, 22:24, 26, 27, 33, 37:39, 42, 43)]) 
```
climaticMoistureIndex has the highest VIF value. Followed by BIO13. 
climaticMoistureIndex = Metric of relative wetness and aridity
BIO13 = Precipitation of Wettest Month
ClimaticMoistureIndex has a lot of information, when looking at the hypothesis. BIO13 is less important, so delete BIO13. 
### Round twenty-one
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32), BIO13 (37)
vif(GIS_woody[,c(3:5, 7, 8, 10, 14, 15, 17, 18, 20, 22:24, 26, 27, 33, 38, 39, 42, 43)]) 
```
BIO3 has the highest VIF value.
BIO3 = Isothermality (BIO2/BIO7) (* 100)
Temperature related, not specifically important for this trait. So okay to delete. 

### Round twenty-two
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32), BIO13 (37), BIO3 (27)
vif(GIS_woody[,c(3:5, 7, 8, 10, 14, 15, 17, 18, 20, 22:24, 26, 33, 38, 39, 42, 43)]) 
```
climaticMoistureIndex has the highest VIF value. Followed by BIO9. 
climaticMoistureIndex = Metric of relative wetness and aridity
BIO9 = Mean Temperature of Driest Quarter
ClimaticMoistureIndex has a lot of information, when looking at the hypothesis. BIO13 is temperature related, not specifically important for this trait, so okay to delete.

### Round twenty-three
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32), BIO13 (37), BIO3 (27), BIO9 (33)
vif(GIS_woody[,c(3:5, 7, 8, 10, 14, 15, 17, 18, 20, 22:24, 26, 38, 39, 42, 43)]) 
```
climaticMoistureIndex has the highest VIF value. Followed by aridityIndexThornthwaite. 
climaticMoistureIndex = Metric of relative wetness and aridity
aridityIndexThornthwaite = Thornthwaite aridity index
ClimaticMoistureIndex has a lot of information, when looking at the hypothesis. aridityIndexThornthwaite contains less information, so okay to delete.

### Round twenty-four
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32), BIO13 (37), BIO3 (27), BIO9 (33), aridityIndexThornthwaite (7)
vif(GIS_woody[,c(3:5, 8, 10, 14, 15, 17, 18, 20, 22:24, 26, 38, 39, 42, 43)]) 
```
climaticMoistureIndex has the highest VIF value. Followed by embergerQ 
climaticMoistureIndex = Metric of relative wetness and aridity
embergerQ = Emberger’s pluviothermic quotient 
ClimaticMoistureIndex has a lot of information, when looking at the hypothesis. embergerQ contains less information, so okay to delete.

### Round twenty-five
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32), BIO13 (37), BIO3 (27), BIO9 (33), aridityIndexThornthwaite (7), embergerQ (10)
vif(GIS_woody[,c(3:5, 8, 14, 15, 17, 18, 20, 22:24, 26, 38, 39, 42, 43)]) 
```
climaticMoistureIndex has the highest VIF value. Followed by BIO2 
climaticMoistureIndex = Metric of relative wetness and aridity
BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp)) 
ClimaticMoistureIndex has a lot of information, when looking at the hypothesis. BIO2 is temperature related, so okay to delete.

### Round twenty-six
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32), BIO13 (37), BIO3 (27), BIO9 (33), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26)
vif(GIS_woody[,c(3:5, 8, 14, 15, 17, 18, 20, 22:24, 38, 39, 42, 43)]) 
```
ClimaticMoistureIndex has the highest VIF value, followed by BIO18, and those two are highly correlated with each other. In my opinion, ClimaticMoistureIndex is important for the model. I will try to do the analysis both with this variable, and without. 

### Round twenty-seven
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32), BIO13 (37), BIO3 (27), BIO9 (33), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), ClimaticMoistureIndex (8)
vif(GIS_woody[,c(3:5, 14, 15, 17, 18, 20, 22:24, 38, 39, 42, 43)])
```
Leaving out ClimaticMoistureIndex results in only VIF values below 10. 

## other options:

### With ClimaticMoistureIndex
To include ClimaticMoistureIndex ánd keep all the VIF values below 10, BIO18, BIO14, and monthCountByTemp10 have to be deleted, resulting in:
  ```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32), BIO13 (37), BIO3 (27), BIO9 (33), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), BIO18 (42), BIO14 (38), and monthCountByTemp10 (15)
vif(GIS_woody[,c(3:5, 8, 14, 17, 18, 20, 22:24, 39, 43)])
```

### With BIO6 (30) 
This increases the VIF value of PETseasonality and BIO6 has a high VIF value as well. But it is not necessary to include this in the analysis, we have three other PET variables. If we  want BIO6 in the analysis, we will have to delete: PETseasonality (18) and ClimaticMoistureIndex (8)
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32), BIO13 (37), BIO3 (27), BIO9 (33), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETseasonality (18) and ClimaticMoistureIndex (8)
vif(GIS_woody[,c(3:5, 14, 15, 17, 20, 22:24, 30, 38, 39, 42, 43)]) 
```

### With growingDegDays0
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), maxTempColdest (13), PETColdestQuarter (16), 
#BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), 
#PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), monthCountByTemp10 (15), BIO14 (38), minTempWarmest (14)
vif(GIS_woody[,c(3:5, 11, 17, 18, 20, 22:24, 39, 42, 43)])
```

### With aridityIndexThornthwaite
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), 
#BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), 
#PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), embergerQ (10), BIO2 (26), BIO14 (38), BIO18 (42)
vif(GIS_woody[,c(3:5, 7, 14, 15, 17, 18, 20, 22:24, 37, 39, 43)])
vif(GIS_woody[,c(3:5, 7, 14, 15, 17, 18, 20, 22:24, 39, 40, 43)])
vif(GIS_woody[,c(3:5, 7, 14, 15, 17, 18, 20, 22:24, 39, 43)])
#added 37, 40
```

### With embergerQ
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), 
#BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), 
#PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), BIO2 (26), BIO14 (38), BIO18 (42), PETseasonality (18)
vif(GIS_woody[,c(3:5, 7, 9, 10, 14, 15, 17, 20, 22:24, 39, 43)])
vif(GIS_woody[,c(3:5, 10, 14, 15, 17, 20, 22:24, 28, 38, 39, 43)])
vif(GIS_woody[,c(3:5, 10, 14, 15, 17, 20, 22:24, 30, 41, 39, 43)])
vif(GIS_woody[,c(3:5, 10, 14, 15, 17, 20, 22:24, 31, 39, 43)])
#added 7, 9, 28, 30, 31, 38, 41
```

### With BIO6 and ClimaticMoistureIndex
Or, if we want to keep both variables: PETseasonality (18), BIO18 (42), monthCountByTemp10 (15), and BIO14 (38) have to be deleted
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32), BIO13 (37), BIO3 (27), BIO9 (33), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETseasonality (18), BIO18 (42), monthCountByTemp10 (15), and BIO14 (38)
vif(GIS_woody[,c(3:5, 8, 14, 17, 20, 22:24, 30, 39, 43)]) 
```

### With BIO6 and aridityIndexThornthwaite
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32), BIO13 (37), BIO3 (27), BIO9 (33), embergerQ (10), BIO2 (26), PETseasonality (18) and ClimaticMoistureIndex (8), BIO14 (38), monthCountByTemp10 (15), BIO18
vif(GIS_woody[,c(3:5, 7, 10, 14, 15, 17, 20, 22:24, 30, 39, 43)]) 
vif(GIS_woody[,c(3:5, 7, 14, 17, 20, 22:24, 30, 37, 39, 43)]) 
vif(GIS_woody[,c(3:5, 7, 14, 17, 20, 22:24, 30, 39, 40, 43)]) 
#added 10, 15, 37, 40
```

### With BIO6 and embergerQ
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), 
#BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), growingDegDays5 (12), climaticMoistureIndex (8), 
#PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), BIO2 (26), BIO14 (38), BIO18 (42), PETseasonality (18)
vif(GIS_woody[,c(3:5, 7, 10, 14, 15, 17, 20, 22:24, 30, 39, 43)])
vif(GIS_woody[,c(3:5, 10, 14, 15, 17, 20, 22:24, 30, 38, 39, 43)])
vif(GIS_woody[,c(3:5, 10, 14, 15, 17, 20, 22:24, 30, 39, 41, 43)])
#added 7, 38, 41
```

### With growingDegDays0 and ClimaticMoistureIndex
```{r}
#excluding: BIO7 (31), BIO11 (35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), maxTempColdestMonth (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), PETWarmestQuarter (19), BIO8 (32), BIO13 (37), BIO3 (27), BIO9 (33), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), BIO18 (42), BIO14 (38), and monthCountByTemp10 (15), minTempWarmest (14)
vif(GIS_woody[,c(3:5, 8, 11, 17, 18, 20, 22:24, 39, 43)])
```

### With growingDegDays0 and aridityIndexThornthwaite
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), maxTempColdest (13), PETColdestQuarter (16), 
#BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), 
#PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), embergerQ (10), BIO2 (26), BIO14 (38), BIO18 (42), monthCountByTemp10 (15), PETDriestQuarter (17), PETseasonality (18)
vif(GIS_woody[,c(3:5, 7, 10, 11, 14, 19, 20, 22:24, 39, 43)])
vif(GIS_woody[,c(3:5, 7, 11, 14, 20, 22:24, 26, 37, 39, 43)])
vif(GIS_woody[,c(3:5, 7, 11, 14, 20, 22:24, 39, 40, 43)])
vif(GIS_woody[,c(3:5, 7, 11, 14, 20, 22:24, 39, 42, 43)])
#added 10, 19, 26, 37, 40, 42
```

### With growingDegDays0 and embergerQ
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), maxTempColdest (13), PETColdestQuarter (16), 
#BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), 
#PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), BIO2 (26), BIO14 (38), BIO18 (42), PETseasonality (18), monthCountByTemp10 (15), minTempWarmest (14)
vif(GIS_woody[,c(3:5, 7, 9, 10, 11, 17, 20, 22:24, 39, 43)])
vif(GIS_woody[,c(3:5, 10, 11, 17, 18, 20, 22:24, 39, 43)])
vif(GIS_woody[,c(3:5, 10, 11, 17, 20, 22:24, 27, 38, 39, 43)])
vif(GIS_woody[,c(3:5, 10, 11, 17, 20, 22:24, 28, 39, 41, 43)])
vif(GIS_woody[,c(3:5, 10, 11, 17, 20, 22:24, 29, 39, 43)])
#added 7, 9, 18, 27, 28, 29, 38, 41
```

## Second selection

### 13. With things
```{r}
vif(GIS_woody[,c(3:5, 14, 15, 17, 18, 20, 22:24, 38, 39, 42, 43)])
```

### 14. With annualPET
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETWettestQuarter (20)
vif(GIS_woody[,c(3:6, 14, 15, 17, 18, 22:24, 38, 39, 42, 43)])
```

### 15. With continentality
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETWettestQuarter (20), PETseasonality (18)
vif(GIS_woody[,c(3:6, 9, 14, 15, 17, 22:24, 38, 39, 42, 43)])
vif(GIS_woody[,c(3:5, 9, 14, 15, 16, 17, 22:24, 38, 39, 42, 43)])
vif(GIS_woody[,c(3:5, 9, 14, 15, 17, 19, 22:24, 38, 39, 42, 43)])
vif(GIS_woody[,c(3:5, 9, 14, 15, 17, 20, 22:24, 38, 39, 42, 43)])
vif(GIS_woody[,c(3:5, 9, 14, 15, 17, 22:24, 26, 38, 39, 42, 43)])
vif(GIS_woody[,c(3:5, 9, 14, 15, 17, 22:24, 27, 38, 39, 42, 43)])
vif(GIS_woody[,c(3:5, 9, 14, 15, 17, 22:24, 32, 38, 39, 42, 43)])
#added 6, 16, 19, 20, 26, 27, 32
```

### 16. With growingDegDays5
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), annualPET (6), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), monthCountByTemp10 (15), bio14 (38), PETseasonality (18)
vif(GIS_woody[,c(3:5, 12, 14, 17, 20, 22:24, 39, 42, 43)])
```

### 17. With maxTempColdest
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), monthCountByTemp10 (15), PETseasonality (18)
vif(GIS_woody[,c(3:5, 13, 14, 17, 20, 22:24, 38, 39, 42, 43)])
#added 38
```

### 18. With PETColdestQuarter
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETseasonality (18), monthCountByTemp10 (15)
vif(GIS_woody[,c(3:5, 14, 16, 17, 20, 22:24, 38, 39, 42, 43)])
#added 38
```

### 19. With PETWarmestQuarter
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETWettestQuarter (20)
vif(GIS_woody[,c(3:5, 14, 15, 17:19, 22:24, 38, 39, 42, 43)])
```

### 20. With thermicityIndex
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), monthCountByTemp10 (15), PETseasonality (18)
vif(GIS_woody[,c(3:5, 14, 17, 20:24, 38, 39, 42, 43)])
#added 38
```

### 21. With bio1
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), monthCountByTemp10 (15), bio14 (38), minTempWarmest (14)
vif(GIS_woody[,c(3:5, 17, 18, 20, 22:24, 25, 38, 39, 42, 43)])
#added 38
```

### 22. With bio2
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETWettestQuarter (20)
vif(GIS_woody[,c(3:5, 14, 15, 17, 18, 22:24, 26, 38, 39, 42, 43)])
```

### 23. With bio3
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETseasonality (18)
vif(GIS_woody[,c(3:5, 14, 15, 17, 20, 22:24, 27, 38, 39, 42, 43)])
```

### 24. With bio4
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), 
#BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), 
#PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETseasonality (18)
vif(GIS_woody[,c(3:5, 14, 15, 17, 20, 22:24, 28, 38, 39, 42, 43)])
```

### With bio5
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), minTempWarmest (14), bio14 (38), PETseasonality (18)
vif(GIS_woody[,c(3:5, 15, 17, 20, 22:24, 26, 29, 38, 39, 42, 43)])
vif(GIS_woody[,c(3:5, 15, 17, 20, 22:24, 29, 30, 39, 42, 43)])
#added 26, 30, 38
```

### With bio7
```{r}
#excluded: BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETseasonality (18)
vif(GIS_woody[,c(3:5, 14, 15, 17, 20, 22:24, 31, 38, 39, 42, 43)])
```

### With bio8 
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETWettestQuarter (20)
vif(GIS_woody[,c(3:5, 14, 15, 17, 18, 22:24, 32, 38, 39, 42, 43)])
```

### With bio9
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETDriestQuarter (17)
vif(GIS_woody[,c(3:5, 14, 15, 18, 20, 22:24, 33, 38, 39, 42, 43)])
```

### With bio10
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), minTempWarmest (14)
vif(GIS_woody[,c(3:5, 15, 17, 18, 20, 22:24, 34, 38, 39, 42, 43)])
```

### With bio11
```{r}
#excluded: BIO7 (31), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), PETseasonality (18), monthCountByTemp10 (15)
vif(GIS_woody[,c(3:5, 14, 17, 20, 22:24, 35, 38, 39, 42, 43)])
#added 38
```

### With bio12
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), bio18 (42), bio14 (38)
vif(GIS_woody[,c(3:5, 14, 15, 17, 18, 20, 22:24, 36, 39, 43)])
```

### With bio13
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), bio18 (42)
vif(GIS_woody[,c(3:5, 14, 15, 17, 18, 20, 22:24, 37:39, 43)])
```

### With bio16
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO12 (36), BIO17 (41), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), bio18 (42)
vif(GIS_woody[,c(3:5, 14, 15, 17, 18, 20, 22:24, 38:40, 43)])
```

### With bio17
```{r}
#excluded: BIO7 (31), BIO11(35), continentality (9), BIO10 (34), BIO4 (28), BIO1 (25), growingDegDays0 (11), maxTempColdest (13), PETColdestQuarter (16), BIO16 (40), BIO12 (36), thermicityIndex (21), BIO5 (29), BIO6 (30), growingDegDays5 (12), climaticMoistureIndex (8), PETWarmestQuarter (19), BIO8 (32), BIO3 (27), BIO9 (33), BIO13 (37), aridityIndexThornthwaite (7), embergerQ (10), BIO2 (26), bio14 (38), bio18 (42)
vif(GIS_woody[,c(3:5, 14, 15, 17, 18, 20, 22:24, 37, 39, 41, 43)])
vif(GIS_woody[,c(3:5, 14, 15, 17, 18, 20, 22:24, 39, 40, 41, 43)])
#added 37, 40
```
